Worflow

### 2017-06-06 R/workflow.Rmd
organize vignette, add summary schema

---
title: "Worflow"
author: "Fanny Perraudeau"
date: "05/25/2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

@@

```{r}
runZinb = TRUE
runClus = TRUE
NCORES = 2
```

@@

- Fit zinbwave with K = 50, X including the batches,   
- Get W which should capture the biology,  
- Use clusterExperiment on W,   
- Get the cluster labels,   
- Use slingshot on W from zinbwave and with cluster labels from clusterExperiment,  
- Get lineages,  
- Fit again zinbwave but with X including batches and lineages from slingshot,  
- Get DE genes between lineages.

Along the worflow, use deviance residuals as adjusted "counts".

@@

# Unnormalized data
I am still using the data Davide gave me a while ago. And I think now it is not the same data Davide is using in its most recent vignette (that is zinb_clustering_over_k.Rmd). It should not be very important but it would be nice if we have a common dataset we can work on.

@@

```{r datain}
#counts<-read.table("../data/oeCufflinkCountData.txt",sep="\t",header=TRUE)
core<-read.table("../data/oeCufflinkCountData_1000Var.txt",sep="\t",header=TRUE)
core<-data.matrix(core)
metadata<-read.table("../data/oeMetadata.txt",sep="\t",header=TRUE)
# load("../data/Expt4c2b_filtdata.Rda")
# load("../data/E4c2b_slingshot_wsforkelly.RData")
```

@@

```{r filtering}
batch <- metadata$Batch
qc <- metadata[,!names(metadata)%in% c("Batch","Experiment","clusterLabels")]
clus.labels<-metadata[,"clusterLabels"]
# vars <- rowVars(log1p(counts))
# names(vars) <- rownames(counts)
# vars <- sort(vars, decreasing = TRUE)
# core <- counts[names(vars)[1:1000],]
```

@@

```{r tabbatch}
table(batch)
```

Cells have been clustered into `r length(unique(clus.labels))` different clusters

```{r tabclus}
table(clus.labels)
```

```{r}
compute.naive.residuals <- function(Y, zinb){
  mu_hat = t(getMu(zinb)) 
  pi_hat = t(getPi(zinb)) 
  Y_hat = (1 - pi_hat) * mu_hat
  Y - Y_hat 
}

compute.pearson.residuals <- function(Y, zinb){
  num = compute.naive.residuals(Y, zinb)
  mu = t(getMu(zinb)) 
  pi = t(getPi(zinb)) 
  phi = matrix(rep(getPhi(zinb), ncol(Y)), ncol = ncol(Y))
  var_hat = (1 - pi) * mu * (1 + mu * (phi + pi))
  num / sqrt(var_hat)  
}

compute.zinb.loglik <- function(Y, zinb){
  mu = t(getMu(zinb)) 
  theta = getTheta(zinb)
  theta_mat = matrix(rep(theta, ncol(Y), ncol = ncol(Y)))
  pi = t(getPi(zinb))
  log( pi * (Y == 0) + (1 - pi) * dnbinom(Y, size = theta, mu = mu) )
}

# 0. Residuals for visualization
## Fit zinbwave X intercept, no batch
Let's run zinbwave with K = 0 and X batches.

```{r}
fn = '../data/zinb_k0_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  print(system.time(zinb0 <- zinbFit(core, K = 0, X = mod, ncores = NCORES)))
  save(zinb0, file = fn)
}else{
  load(fn)
}
```

We use deviance residuals for visualization. Let's check that deviance residuals look ok.

```{r}
res = compute.deviance.residuals(core, zinb0)
res[1:3,1:3]
```

PCA on the residuals where color are for batches on the left and previously found clusters on the right.

```{r}
pca = prcomp(t(res))
par(mfrow = c(1,2))
plot(pca$x, col = batch, pch = 19,
     main="PCA of residuals\ncolor=batch")
plot(pca$x, col = clus.labels, pch = 19,
     main = "PCA of residuals\ncolor=cluster")
par(mfrow = c(1,1))
```

Boxplot of the residuals for each cell.

```{r}
res_order = res[, order(as.numeric(batch))]
col_order = as.numeric(batch)[order(as.numeric(batch))]
boxplot(res_order, main='Boxplot of residuals\ncolor=batch', col = col_order,         staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
```

Deviance residuals seem ok.

# 1. zinbwave with raw counts

Let's run zinbwave with K = 50 and X including batches.

```{r}
fn = '../data/zinb_k50_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50,
                                      X = mod)))
  save(zinb50, file = fn)
}else{
  load(fn)
}
```

# 2. clusterExperiment with W

@@

```{r rsec_50}
fn = '../data/clustExp_W_k50.rda'
metadata$clusterLabels[is.na(metadata$clusterLabels)]<- -2 #symbol for samples missing from original clustering
seObj<-SummarizedExperiment(t(W),colData=metadata)
if(runClus) {
 print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1), betas = 0.8,
                      clusterFunction = "hierarchical01", minSizes=1,
                      ncores = NCORES,
					  combineProportion = 0.7,mergeMethod = "locfdr", mergeCutoff = 0.01
                      subsampleArgs = list(resamp.num=100,
                                           clusterFunction="kmeans",
                                           clusterArgs=list(nstart=10)),
                      seqArgs = list(k.min=3, top.can=5), verbose=TRUE))
save(ceObj, file = fn)
}
else{
	load(fn)
}
```

```{r examineCombineMany}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(100)))
plotCoClustering(ceObj)
table(primaryClusterNamed(ceObj))
sum(primaryCluster(ceObj) == -1)
```

@@

```{r compareClusters}
plot(getW(zinb50), col=clus.labels, main = 'color = previous')
plot(getW(zinb50), col=as.factor(primaryCluster(ceObj)), main = 'color = new')
table(data.frame(previous = clus.labels, new = primaryCluster(ceObj)))
```

## Original ClusterExperiment code (in Steps)
I use the code of Davide from zinb_clutering_over_k.Rmd

```{r consensus_50, eval=FALSE}
## matching the parameters used by Russell
W = zinb50@W
fn = '../data/clustExp_W_k50.rda'
if(runClus) {
  cl_res <- clusterMany(t(W), ks = 4:15, alphas = c(0.1), betas = 0.8,
                        clusterFunction = "hierarchical01", minSizes=1,
                        sequential = TRUE, subsample = TRUE, ncores = 7,
                        subsampleArgs = list(resamp.num=100,
                                             clusterFunction="kmeans",
                                             clusterArgs=list(nstart=10)),
                        seqArgs = list(k.min=3, top.can=5), verbose=TRUE)
  save(cl_res, file = fn)
} else {
  load(fn)
}
```

```{r combine_50, eval=FALSE}
combined <- combineMany(cl_res, proportion = 0.7)
plotClusters(combined, colPalette = c(bigPalette, rainbow(100)))
plotCoClustering(combined)
table(primaryClusterNamed(combined))
sum(primaryCluster(combined) == -1)
```

```{r new_clusters, eval=FALSE}
plot(getW(zinb50), col=clus.labels, main = 'color = previous')
plot(getW(zinb50), col=as.factor(primaryCluster(combined)), main = 'color = new')
table(data.frame(previous = clus.labels, new = primaryCluster(combined)))
```

```{r merged, eval=FALSE}
combined <- makeDendrogram(combined)
plotDendrogram(combined)
merged <- mergeClusters(combined, mergeMethod = "locfdr", cutoff = 0.01)
```

Just to see what it looks like, let's look at the heatmap on the W (number of cells x K=50). Now, we would like to plot a heatmap using the residuals (number of cells x number of genes). When I call plotHeatmap using argument visualizeData = residuals_matrix, it does not work and says that if give separate visualizeData, must be of same dimensions as assay(data).

```{r heatmapOld, eval=FALSE}
plotHeatmap(merged, clusterSamplesData="dendrogramValue", breaks = .99)
```

@@

heatmap.2(res, Colv = merged@dendro_samples,
          scale = 'none', trace = 'none', 
          col = colorRampPalette(c('blue', 'yellow'))(51),
          ColSideColors = pal[primaryClusterNamed(merged)],
          labCol = '', main = 'Deviance Residuals, 1000 most variable genes')
legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
       horiz = T, cex = 0.5)
```

# 3. slingshot with W and cluster labels
I used the unsupervised mode on W got from fitting zinbwave with batches in X and K = 4.

@@

# 4. zinbwave with lineages

Which clusters would it make sense to test one versus the other? Would it be interesting to do two by two comparison between clusters next to each other in the different lineages?

# 5. DE genes with betas

>>>>

---
title: "Worflow"
author: "Workflow team"
date: "06/06/2017"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

@@

```{r}
runZinb <- FALSE
runClus <- TRUE
NCORES <- 8
```

FP: yes, I agree.

```{r residuals}
compute.zinb.loglik <- function(Y, zinb){
  mu = t(getMu(zinb)) 
  theta = getTheta(zinb)
  theta_mat = matrix(rep(theta, ncol(Y), ncol = ncol(Y)))
  pi = t(getPi(zinb))
  log( pi * (Y == 0) + (1 - pi) * dnbinom(Y, size = theta, mu = mu) )
}

@@

- Dimensionality reduction using zinbwave to get W which should capture the biology,
- Cluster cells using clusterExperiment on W to get the cluster labels,   
- Get lineage using slingshot on W and cluster labels from clusterExperiment,  
- Get DE genes between lineages/clusters.

Along the worflow, use deviance residuals as adjusted values.

@@

# Steps of the workflow
## Create a SummarizedExperiment object

Along the workflow, we want to use a unique SummarizedExperiement object carrying all the data we need.

@@

```{r datain}
#counts<-read.table("../data/oeCufflinkCountData.txt",sep="\t",header=TRUE)
core <- read.table("../data/oeCufflinkCountData_1000Var.txt",
                   sep = "\t", header = TRUE)
core <- data.matrix(core)
metadata <- read.table("../data/oeMetadata.txt", sep = "\t", header = TRUE)
# symbol for samples missing from original clustering
metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2
```

@@

```{r batches}
batch <- metadata$Batch
```

```{r}
col_batch = rep(brewer.pal(9, "Set1"), 2)
names(col_batch) = unique(batch)
table(batch)
```

We have qc measures from the data

```{r qc}
qc <- metadata[, !names(metadata) %in% c("Batch", "Experiment", "clusterLabels")]
head(qc, 2)
```

```{r clusterlabels}
clus.labels <- metadata[, "clusterLabels"]
```

In published work (FP: add ref), cells have been clustered into `r length(unique(clus.labels))` different clusters

```{r}
col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
col_clus <- col_clus[1:length(unique(clus.labels))]
names(col_clus) <- sort(unique(clus.labels))
table(clus.labels)
```

@@

Let's create a SummarizedExperiment object to store the raw counts and information about the data, that is batches, published labels, and quality control measures.

```{r}
se <- SummarizedExperiment(assays = list(rawCounts = core), colData = metadata)
```

## Dimensionality reduction adjusting for gene and cell level covariates

To cluster and get lineages we want to reduce the dimension of the data. We are going to use zinbwave to do so. First, let's fit zinbwave with first K = 0 to compute normalized values (i.e. deviance residuals) adjusted for batches. We could also adjust for gene length or GC content here. We then fit zinbwave to get the dimensionality reduced matrix W specifying the number of dimension K = 50. Eventually, we will call zinbwave just once where we would have an argument in zinbFit like "compute_normalized_values" in c(TRUE, FALSE). For K = 0 and K = 50, we correct for batch effect including batches in X.

```{r}
fn50 <- '../data/zinb_k50_batch.rda'
fn0 <- '../data/zinb_k0_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  # K = 0
  print(system.time(zinb0 <- zinbFit(core, ncores = NCORES, K = 0, X = mod)))
  save(zinb0, file = f50)
  # K = 50
  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50, X = mod)))
  save(zinb50, file = fn50)
}else{
  load(fn50)
  load(fn0)
}
```

### Normalized values

We use deviance residuals as normalized values for visualization. FP: explain rational: K=0 so residuals capture the bio adjusting for batch. Let's check that deviance residuals look ok.

```{r}
res <- compute.deviance.residuals(core, zinb0)
res[1:3,1:3]
```

Boxplot of the normalized values for each cell. It seems that correction for batches is ok.

```{r}
res_order <- res[, order(as.numeric(batch))]
col_order <- as.numeric(batch)[order(as.numeric(batch))]
boxplot(res_order, main='Boxplot of normalized values\ncolor=batch',
        col = col_order, staplewex = 0, outline = 0, border = col_order, xaxt = 'n')
```

PCA on the normalized values where color are for batches on the left and previously found clusters on the right. We want no clustering on the left side and clustering on the right side.

```{r}
pca <- prcomp(t(res))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20,
     main="PCA of normalized values\ncolor=batch")
plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20,
     main = "PCA of normalized values\ncolor=cluster")
par(mfrow = c(1,1))
```

Let's add the normalized values as an assay dataset in our SummarizedExperiment object.

```{r}
assays(se)[['normalizedValues']] <- res
```

### W

Let's check that performing MDS on W we have something coherent with published clusters.

```{r}
W <- getW(zinb50)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(clus.labels)], main = 'MDS', pch = 20,
     xlab = 'Component 1', ylab = 'Component 2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

Let's add W to the colData of our SummarizedExperiment object.

```{r}
W <- data.frame(W)
colnames(W) <- paste0('W', 1:ncol(W))
colData(se) <- cbind(colData(se), W)
```

FP: what do you think of implemented a "compute_normalized_values" argument to zinbFit for summarizedExperiment objects where zinbFit would take as input a SummarizedExperiment object and return a summarizedExperiment object with slots added for normalized values and W?

## Cluster of cells

We use clusterExperiment with W.

@@

FP: What about the following call to RSEC?

```{r rsec_50}
W = colData(se)[, grepl('^W', colnames(colData(se)))]
W = as.matrix(W)
fn = '../data/RSEC_W.rda'
if (runClus){
  print(system.time(ceObj <- RSEC(t(W), k0s = 4:15, 
                                  clusterFunction = "hierarchical01",
                                  minSizes = 5,
                                  ncores = NCORES,
                                  combineProportion = 0.7,
                                  mergeMethod = "locfdr", mergeCutoff = 0.01)))
  save(ceObj, file = fn)
}else{
  load(fn)
}
```

```{r examineCombineMany}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(100)))
```

```{r plotcoclust}
plotCoClustering(ceObj)
```

```{r}
table(primaryClusterNamed(ceObj))
sum(primaryCluster(ceObj) == -1)
```

@@

```{r compareClusters}
plot(W, col = col_clus[as.character(clus.labels)], main = 'color = previous', pch = 20)
plot(W, col = as.factor(primaryCluster(ceObj)), main = 'color = new', pch = 20)
table(data.frame(previous = clus.labels, new = primaryCluster(ceObj)))
```

@@

plotHeatmap(res, Colv = merged@dendro_samples,
          scale = 'none', trace = 'none', 
          col = colorRampPalette(c('blue', 'yellow'))(51),
          ColSideColors = pal[primaryClusterNamed(merged)],
          labCol = '', main = 'Deviance Residuals, 1000 most variable genes')
legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
       horiz = T, cex = 0.5)
```

## Get lineages
I used the unsupervised mode on W got from fitting zinbwave with batches in X and K = 4.

@@

## DE analysis
